### 2021-08-03 paper/paper.tex

We are broadly interested in simulation-focused scientific computing, in particular
numerical approximations for hyperbolic partial differential equations (PDEs), computational fluid dynamics (CFD),
and related problems. The focus of our research ranges from specific applications
in CFD over general multi-physics coupling strategies to the required high-order
numerical methods themselves. In addition, we are academics and involved in teaching
students in these areas of science. Thus, we would like to have a code that is
\begin{enumerate}
  \item extendable to be useful for research and development
  \item easy to apply/understand to be useful for students/collaborators
  \item fast enough to be useful not only for small toy problems
\end{enumerate}
Additionally, we are greedy\footnote{See \url{https://julialang.org/blog/2012/02/why-we-created-julia/}} and want to have all these features in a single code.
Roughly one year of collaborative work has resulted in the current version of
\trixi\footnote{https://github.com/trixi-framework/Trixi.jl}, which provides adaptive
high-order numerical simulations of hyperbolic partial differential equations (PDEs)
in Julia \cite{bezanson2017julia}. Starting as an experiment, we have been able
to reach more and more of our goals with \trixi.

\todo{TODO: Jesse wants to write about the \trixi community} %TODO: Jesse

@@

The design of \trixi was guided by extensibility, ease of use, and efficiency
(in this particular order). Thus, it will likely be possible to make (some parts)
even faster while sacrificing simplicity. Nevertheless, \trixi is already quite
fast, even compared to established high-performance open source software,
as described in detail in Section~\ref{sec:performance-comparison}. In this
section, we focus mostly on the first two goals.

@@

The most important design decision was to create \trixi as a library. In the
context of simulation-focused scientific computing, there are many (open
or closed source) codes designed as monolithic applications that can
simulate a specific setup. Such monolithic codes can often be configured in limited amounts by
compile time options and parameter files. In contrast, users of \trixi do not
need to compile anything or learn the syntax of parameter files. They just load
the library and write their simulation setups in Julia. Such scripts describing
a simulation setup for \trixi are called elixirs. As of version v0.3.50, \trixi
comes with \ca 200 example elixirs. A basic elixir could look as follows.
% TODO: Update Trixi version above

@@

In traditional monolithic code bases, it is often necessary to implement a new
feature for all combinations of possible (compile time) options. This effort
can make it difficult to experiment with new ideas. In contrast, Julia's dynamic
nature, multiple dispatch, and just-ahead-of-time compilation allow us to implement only the features
that are strictly necessary. For example, if a user wants to simulate a new
physics model only on Cartesian grids with smooth solutions, there is no need
to implement anything for curvilinear coordinates or shock capturing approaches.
In addition, it allows a user to extend \trixi step by step with new capabilities.
For example, there are ongoing efforts to incorporate summation by parts finite
difference methods via SummationByPartsOperators.jl \cite{ranocha2021sbp} and
more discontinuous Galerkin methods on simplex elements via StartUpDG.jl\footnote{\url{https://github.com/jlchan/StartUpDG.jl}} in \trixi.
This is feasible because the modular design of \trixi gives developers the flexibility to use
a subset of the available meshes/methods, selected via multiple dispatch,
to test their newly implemented feature(s).

@@

Note that we do not want to state that Julia is generically faster than Fortran,
C, or C++. In contrast, we would like to emphasize that Julia can be at least
as fast as code written in these traditional scientific computing languages.
We believe that it will be possible to get the same efficiency by rewriting
FLUXO appropriately. However, the code introspection and profiling tools available
in Julia make it easier for us to optimize Julia code compared to other languages
we are familiar with (having a diverse team with a strong background in C, C++,
and Fortran).

@@

Julia was designed from scratch to be useful for numerical computing
\cite{bezanson2017julia}. Its usefulness has been demonstrated for example for
GPU programming \cite{besard2018juliagpu, omlin2020solving},
optimization \cite{dunning2017jump}, data science including
big data\footnote{Celeste project: \url{https://github.com/jeff-regier/Celeste.jl},
see \url{https://juliacomputing.com/case-studies/celeste/} (accessed 2021-07-22)},
machine learning \cite{innes2018fashionable}, and scientific machine learning
(SciML) \cite{pal2021opening}. There are also mature libraries for ODE problems
\cite{rackauckas2017differentialequations}, small scale spectral approximations
\cite{olver2014practical}, and classical finite element methods \cite{badia2020gridap}.
In addition, there are some Julia packages dedicated to solve specific time-dependent
PDEs such as \cite{ramadhan2020oceananigans, constantinou2021geophysicalflows}.
However, there does not seem to be a general framework of high-order methods
or hyperbolic PDEs in Julia. Thus, \trixi is an ideal candidate for a case study
of Julia for simulation-focused scientific computing, an area where codes written
in classical programming languages such as C and Fortran are still dominant
\cite{krais2021flexi, parsani2021ssdc}.

@@

Julia itself and most packages are developed on GitHub. The community provides
many tools to use corresponding CI setups to prepare and publish
documentation\footnote{\url{https://github.com/JuliaDocs/Documenter.jl}},
and related tasks. This setup also encourages other good software development
practices that we use in \trixi such as mandatory code review before pull requests
are merged.

@@

Due to Julia's package manager, the cost of using external dependencies is low
enough to not be a problem. Traditionally, many scientific simulation codes tend
to reduce the number of external dependencies as much as possible due to the
complexity of handling different versions and making everything work together.
Such an approach often leads to a lot of code duplication. For example, many
CFD codes implement their own time integration methods, arguing that the spatial
part is much more complex. While this is often true, it is still more efficient
in terms of human workload to be able to reuse existing implementations. This
allows experts on time integration methods to develop specialized algorithms and
implement them in open source software while practitioners or researchers focusing
on spatial semidiscretizations ca benefit immediately by changing only one line
of code. For example, optimized time integration methods were developed in
\cite{ranocha2021optimized} and implemented in OrdinaryDiffEq.jl. Thus, they
can be used with \trixi by changing a single line of code in an elixir file. In contrast, researchers
working without external dependencies would need to use a day or two to digest
all the details and implement such time integration methods in their own codes, making it much
less likely that they will do this and benefit from recent algorithmic developments.

@@

Such a difference between the programming languages used in the front- and
back-ends lead to a well-known barrier between a user and the developers of a library.
Julia was designed from the beginning to make it possible to write both simple
high-level code and highly efficient compute kernels, thereby solving the
two-language problem --- at least to a large extent. Highly efficient low-level
code will of course still look different from simple high-level code, but the
barrier between them is much smaller. In particular, this makes it easier for
users to contribute to a package and act as developers. This also contributes
to the following observation.

\subsubsection{The code base is simple enough to be useful for students}

@@

Julia itself is released under the MIT license and ships some GPL-licensed
third-party software (that can optionally be disabled for commercial purposes).
Many packages in the Julia ecosystem follow this approach and are freely available
under the permissive MIT license, including \trixi. This allows using Julia
without the need to pay for commercial software. In particular, students can
work with the software on their private computers without restrictions.

@@

Based on our experience, we consider Julia to be suitable for simulation-focused
scientific computing, in particular for hyperbolic PDEs, computational fluid
dynamics, and related problems --- at least on the scale of shared memory
parallelism. The scalability of high-order methods for hyperbolic PDEs written
in Julia to high performance computing applications still needs to be demonstrated.
However, we do not expect unsolvable problems there, as Julia is not generically
slower than traditional compiled HPC programming languages. Nevertheless, it
appears to be more complicated to scale Julia to HPC while still keeping the
code as simple, clean, and flexible as it is now.

Learning a new programming language requires of course some effort and needs some
time. Nevertheless, it pays off in our experience and we do not want to miss the
nice features of \trixi enabled by Julia and its ecosystem. Thus, while there is
of course no need to switch to Julia if people are satisfied with their existing
tools, we encourage researchers to try out Julia and stay for a while. The Julia
community (on the Julia Discourse forum\footnote{\url{https://discourse.julialang.org/}}
or Slack/Zulip workspaces) is usually welcoming and helpful, both for new and
experienced users. Making use of existing learning material and asking questions
appears to be important since a new programming style is necessary to make full
use of all of Julia's strengths.

@@

% References
\input{bib.tex}

\todo{ % TODO
\begin{itemize}
  \item Full paper: 5--10 pages
  \item Prepare repro repo including scripts for \trixi and FLUXO
  \item Restructure non-conservative stuff and add MHD benchmarks
  \item Shall we re-create the schematic overview in tikz?
  \item Shall we compare \trixi (in terms of structure, not performance!)
        to other open source frameworks such as MFEM,
        NGSolve, DealII, DUNE, FENICS, Firedrake, ClawPack, ...? Definitely not
        explicitly to not make people angry, but maybe implicitly without
        mentioning their names?
        \begin{itemize}
          \item Gregor will think about it!
        \end{itemize}
  \item Are there other aspects we should discuss?
  \item Check affiliations
  \item Check order of authors
  \item Maybe include some cool pictures? Give a flavor of the kind of simulations
        you can do with \trixi? Pictures trigger people! For example after the
        list of main features?
  \item Summary of community, collaborators, and users: GitHub statistics, contributors,
        Vikram and Chandrashekar, Steven Frankel and Philipp Öffner (maybe), ...
        Related to the relatively large number of students!
        $\to$ Jesse will make a PR!
  \item A simple example where we really use multiple dispatch?
        \begin{itemize}
          \item Just pass functions around
          \item Dispatch on parameters of DG struct
        \end{itemize}
\end{itemize}
\begin{quote}
  Compared to an extended abstract, a full paper presents more of the background
  and context motivating the work. It compares the work to other approaches taken
  in the field and gives some additional insights on the conference contribution.
  Use cases back up the work by showing how it can be used.
\end{quote}}

>>>>

We are broadly interested in simulation-focused scientific computing, in particular
numerical approximations for hyperbolic partial differential equations (PDEs), computational fluid dynamics (CFD),
and related problems. The focus of our research ranges from specific applications
in CFD to general multi-physics coupling strategies to the high-order numerical methods 
such simulations are built upon. In addition, we are academics and involved in teaching 
students in these areas of science. Thus, we would like to have a 
code that is
\begin{enumerate}
  \item extendable to be useful for research and development
  \item easy to apply/understand to be useful for students/collaborators
  \item fast enough to be useful beyond small toy problems
\end{enumerate}
Additionally, we are greedy\footnote{See \url{https://julialang.org/blog/2012/02/why-we-created-julia/}} 
and wish to include all these features within a single code.
Roughly one year of collaborative work has resulted in the current version of
\trixi\footnote{\url{https://github.com/trixi-framework/Trixi.jl}}, providing adaptive
high-order numerical simulations of hyperbolic partial differential equations (PDEs)
in Julia \cite{bezanson2017julia}. Starting as an experiment, we have been able
to reach more and more of our goals with \trixi.

@@

The design of \trixi was guided by extensibility, ease of use, and efficiency
(in this particular order). Thus, it is likely possible to make (some parts) even 
faster at the cost of simplicity and readibility. Nevertheless, \trixi is already quite
fast, even compared to established high-performance open source software,
as described in detail in Section~\ref{sec:performance-comparison}. In this
section, we focus mostly on the first two goals.

@@

The most important design decision was to create \trixi as a library. In the
context of simulation-focused scientific computing, there are many (open
or closed source) codes that are designed as monolithic applications that can
simulate a specific setup. These codes can often be configured to a limited extent 
by specifying compile time options or editing parameter files. In contrast, users of 
\trixi do not need to compile anything or learn parameter file syntax. Instead, they can
just load the library and write their simulation setups in Julia. Scripts which describe
a simulation setup for \trixi are referred to as ``elixirs''. As of version v0.3.50, \trixi
comes with \ca 200 example elixirs. A basic elixir might look as follows:
% TODO: Update Trixi version above

@@

In traditional monolithic code bases, it is often necessary to implement a new
feature for all combinations of possible (compile time) options. This effort
can make it difficult to experiment with new ideas. In contrast, Julia's dynamic
nature, multiple dispatch, and just-ahead-of-time compilation allow us to implement only the features
that are strictly necessary. For example, if a user wants to simulate a new
physics model only on Cartesian grids with smooth solutions, there is no need
to implement anything for curvilinear coordinates or shock capturing approaches.
In addition, it allows a user to extend \trixi step by step with new capabilities.
For example, there are ongoing efforts to incorporate summation by parts finite
difference methods via SummationByPartsOperators.jl \cite{ranocha2021sbp} and
discontinuous Galerkin methods on simplex elements via StartUpDG.jl\footnote{\url{https://github.com/jlchan/StartUpDG.jl}} in \trixi.
This is feasible because the modular design of \trixi gives users the flexibility to select
a subset of the available meshes/methods, selected via multiple dispatch,
to test their newly implemented feature(s).

@@

Note that we do not want to state that Julia is generically faster than Fortran,
C, or C++. In contrast, we would like to emphasize that Julia can be at least
as fast as code written in these traditional scientific computing languages.
%We believe that it will be possible to get the same efficiency by rewriting
%FLUXO appropriately. 
Together, the members of our team have strong backgrounds in C, C++, and 
Fortran. Nevertheless, our Julia code is faster than the established HPC Fortran 
code FLUXO for this example. The reason is not that Fortran or Julia are generically 
faster; in contrast, we expect to be able to achieve similar performance from either 
language. Instead, \trixi owes its performance optimizations in part to the code 
introspection and profiling tools available in Julia. Similar tools used to optimize the 
performance in other languages are usually not as easy to use as their Julia 
equivalents (if they exist at all).

@@

Julia was designed from scratch to be useful for numerical computing
\cite{bezanson2017julia}. Its usefulness has been demonstrated for example for
GPU programming \cite{besard2018juliagpu, omlin2020solving},
optimization \cite{dunning2017jump}, data science including
big data\footnote{Celeste project: \url{https://github.com/jeff-regier/Celeste.jl},
see \url{https://juliacomputing.com/case-studies/celeste/} (accessed 2021-07-22)},
machine learning \cite{innes2018fashionable}, and scientific machine learning
(SciML) \cite{pal2021opening}. There are also mature libraries for ODE problems
\cite{rackauckas2017differentialequations}, small scale spectral approximations
\cite{olver2014practical}, and classical finite element methods \cite{badia2020gridap}.
In addition, there are some packages dedicated to solving specific time-dependent
PDEs such as \cite{ramadhan2020oceananigans, constantinou2021geophysicalflows}.
However, there does not seem to be a general framework for high-order methods
for hyperbolic PDEs in Julia. Thus, \trixi is an ideal candidate for a case study
of Julia for simulation-focused scientific computing, an area where codes written
in classical programming languages such as C and Fortran are still dominant
\cite{krais2021flexi, parsani2021ssdc}.

@@

Julia itself and most packages are developed on GitHub. The community provides
many tools to use corresponding CI setups to prepare and publish
documentation\footnote{\url{https://github.com/JuliaDocs/Documenter.jl}},
and related tasks. This setup also encourages other good software development
practices that we use in \trixi, such as mandatory code review before pull requests
are merged.

@@

Due to Julia's package manager, the cost of using external dependencies is low
enough to not be a problem. Traditionally, many scientific simulation codes tend
to reduce the number of external dependencies as much as possible due to the
complexity of handling different versions and making everything work together.
Such an approach typically leads to significant code duplication. For example, many
CFD codes implement their own time integration methods, arguing that the spatial
part is much more complex. While this is often true, it is still more efficient
in terms of human workload to be able to reuse existing implementations. This
allows experts on time integration methods to develop specialized algorithms and
implement them in open source software while practitioners or researchers focusing
on spatial semidiscretizations ca benefit immediately by changing only one line
of code. For example, optimized time integration methods were developed in
\cite{ranocha2021optimized} and implemented in OrdinaryDiffEq.jl. These methods
can be used with \trixi by changing a single line of code. In contrast, researchers
working without external dependencies would need to spend a day or two to digest
details before even beginning implementation. This additional overhead in developer
time makes it less likely that researchers would implement these methods in their own 
codes and benefit from recent algorithmic developments.

@@

Such a difference between the programming languages used in the front- and
back-ends lead to a well-known barrier between a user and the developers of a library.
Julia was designed from the beginning to make it possible to write both simple
high-level code and highly efficient compute kernels, thereby solving the
two-language problem --- at least to a significant extent. Highly efficient low-level
code will of course still look different from simple high-level code, but the
barrier between them is much smaller. In particular, this makes it easier for
users to transition to developers and contribute to a package. This also contributes
to the following observation:

\subsubsection{The code base is simple enough to be useful for new users}

@@

Julia itself is released under the MIT license and ships some GPL-licensed
third-party software (that can optionally be disabled for commercial purposes).
Many packages in the Julia ecosystem follow this approach and are freely available
under the permissive MIT license, including \trixi. This allows programmers to use Julia
without needing to pay for commercial software. In particular, students can
work with the software on their private computers without restrictions.

@@

Based on our experience, we consider Julia to be suitable for simulation-focused
scientific computing, in particular for hyperbolic PDEs, computational fluid
dynamics, and related problems --- at least on the scale of shared memory
parallelism. The scalability of high-order methods for hyperbolic PDEs written
in Julia to high performance computing applications still needs to be demonstrated.
However, we do not expect this to remain an unsolvable problem since Julia is not 
generically slower than traditional compiled HPC programming languages. Nevertheless, 
it appears to be more complicated to scale Julia to HPC while still keeping the
code as simple, clean, and flexible as it is now.

Learning a new programming language requires of course some effort and needs some
time. Nevertheless, it has paid off in our experience and we do not want to miss the
nice features of \trixi enabled by Julia and its ecosystem. Thus, while there is
of course no need to switch to Julia if people are satisfied with their existing
tools, we encourage researchers to try out Julia and stay for a while. The Julia
community (on the Julia Discourse forum\footnote{\url{https://discourse.julialang.org/}}
or Slack/Zulip workspaces) is usually welcoming and helpful, both for new and
experienced users. Making use of existing learning material and asking questions
appears to be important since a new programming style is necessary to make full
use of all of Julia's strengths.

@@

Jesse Chan acknowledges support from the US National Science Foundation under 
awards DMS-1719818 and DMS-1943186. 
% References
\input{bib.tex}

\todo{ % TODO
\begin{itemize}
  \item Full paper: 5--10 pages
  \item Prepare repro repo including scripts for \trixi and FLUXO
  \item Restructure non-conservative stuff and add MHD benchmarks
  \item Shall we re-create the schematic overview in tikz?
  \item Shall we compare \trixi (in terms of structure, not performance!)
        to other open source frameworks such as MFEM,
        NGSolve, DealII, DUNE, FENICS, Firedrake, ClawPack, ...? Definitely not
        explicitly to not make people angry, but maybe implicitly without
        mentioning their names?
        \begin{itemize}
          \item Gregor will think about it!
        \end{itemize}
  \item Are there other aspects we should discuss?
  \item Check affiliations
  \item Check order of authors
  \item Maybe include some cool pictures? Give a flavor of the kind of simulations
        you can do with \trixi? Pictures trigger people! For example after the
        list of main features?
  \item A simple example where we really use multiple dispatch?
        \begin{itemize}
          \item Just pass functions around
          \item Dispatch on parameters of DG struct
        \end{itemize}
\end{itemize}
\begin{quote}
  Compared to an extended abstract, a full paper presents more of the background
  and context motivating the work. It compares the work to other approaches taken
  in the field and gives some additional insights on the conference contribution.
  Use cases back up the work by showing how it can be used.
\end{quote}}
